---
title: "NCA Timber: Lake States"
output: html_notebook
---

```{r}
library(tidyverse)
```


```{r, results = 'hide', message=FALSE, warning=FALSE, echo=FALSE}
# read in .xsl data as dataframe
library(readxl)

# read tables from excel and drop first column

merchBio <- read_excel("Merch Bio GLakes 06-14-2021.xlsx") %>% select(-1)
premerchBio <- read_excel("Premerch Bio GLakes 06-14-2024.xlsx") %>% select(-1)

# load price data
tmnPrices <- read_excel("TMN_Price_Series_September2023.xlsx", sheet = 1) %>%
  filter(Market == 'Stumpage')
```
# Organize the price data

```{r}
# print the list of unique species in alphabetical order
tmnPrices %>% select(Species) %>% unique() %>% arrange(Species)

```
```{r}
# print the list of unique species in alphabetical order from merchBIO using SPGRPNM
merchBio %>% select(SPGRPNM) %>% unique() %>% arrange(SPGRPNM)
```
```{r}
# create a crosswalk between tmnPrices$Species and merchBio$SPGRPNM
# this will allow us to join
crosswalk <- tribble(
  ~SPGRPNM, ~Species,
  'Ash', 'Ash',
  'Basswood', 'Basswood',
  'Beech', 'Beech',
  'Black Walnut', 'Black Walnut',
  'Cottonwood and aspen (East)', 'Aspen', 
  'Eastern hemlock', 'Hemlock', 
  'Eastern white and red pines', 'White Pine',
  'Eastern white and red pines', 'Red Pine',
  'Hard maple', 'Maple',
  'Hickory', 'Hickory',
  'Jack pine', 'Jack Pine',
  'Loblolly and shortleaf pines', '',
  'Other eastern hard hardwoods', '',
  'Other eastern soft hardwoods', '',
  'Other eastern softwoods', '',
  'Other red oaks', '',
  'Other white oaks', '',
  'Select red oaks', 'Red Oak',
  'Select white oaks', 'White Oak',
  'Soft maple', 'Soft Maple',
  'Spruce and balsam fir', 'Spruce',
  'Spruce and balsam fir', 'Spruce/Fir',
  'Tupelo and blackgum', '',
  'Yellow-poplar', '',
  'Yellow birch', 'Yellow Birch')
  

```



# Organize the FIA data

```{r}
# join the two dataframes
df <- left_join(merchBio, premerchBio, by = c('STATENM','STATECD', 'COUNTYNM',
                                              'COUNTYCD', 'SPGRPCD', 'SPGRPNM',
                                              'SPCLASS', 'UNITCD', 'EVALID',
                                              'RESERVCD')) %>%
                  filter(RESERVCD == 0)

# pivot longer on columns 11 to 33
# drop if biomass is NA
df <- df %>% pivot_longer(cols = 11:33,
                          names_to = 'sizeClass',
                          values_to = 'biomass') %>%
            filter(!is.na(biomass))

```

```{r}
# join prices to df
df <- left_join(df, tmnPrices, by = c('SPGRPCD' = 'Species'))
```


# summarize the data

```{r}
# create a table that shows the distribution of biomass by size class and state
df %>% group_by(STATENM, sizeClass) %>%
  summarise(totalBiomass = sum(biomass), .groups = 'drop') %>%
  pivot_wider(names_from = sizeClass, values_from = totalBiomass) %>%
  mutate(total = rowSums(select(., -STATENM))) %>%
  arrange(desc(total))
```

```{r}
# create a panel of histograms by state showing the distribution of biomass,
# restrict the data to drop biomass in the top 10% of the distribution
df %>% group_by(STATENM) %>%
  filter(biomass < quantile(biomass, 0.9)) %>%
  ggplot(aes(x = biomass)) +
  geom_histogram(bins = 30) +
  facet_wrap(~STATENM, scales = 'free_y') +
  labs(title = 'Distribution of Biomass by State',
       x = 'Biomass (kg/ha)',
       y = 'Count')


```


Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

